\textbf{Machine learning classification by fitting amplicon sequences to existing OTUs}

### 2023-04-27 submission/manuscript.Rmd
final revisions

**observation format** (max 1200 words, 2 figures, 25 ref)

@@

There is increasing evidence for an association between the composition of the gut microbiome and a variety of diseases, such as crohn's disease and colorectal cancer [@Morgan2012;@Sobhani2011]. There is great potential diagnose of disease with gut microbiome sequence data and machine learning. Taxonomic composition of microbial
communities can be assessed using amplicon sequencing of the 16S rRNA
gene, which is the input to classification models. Analysis of 16S rRNA
gene sequence data generally relies on assigning sequences into
operational taxonomic units (OTUs). The process of OTU clustering can
either be reference-based or *de novo*. The quality of OTUs generated
with reference-based clustering is generally poor compared to those
generated with *de novo* clustering @westcott2015. While *de novo* clustering produces high-quality OTU clusters where sequences are accurately grouped based on similarity thresholds, the resulting OTU clusters depend on the sequences within the dataset and the addition of new data has the potential to redefine OTU cluster composition. The unstable nature of *de novo* OTU clustering complicates deployment of
machine learning models since integration of additional data requires
reclustering all the data and retraining the model. The ability to
integrate new data into a validated model without reclustering and
retraining could allow for the application of a single model that can
continually classify new data. Recently, Sovacool *et al.* introduced
OptiFit, a method for fitting new sequence data into existing OTUs
@sovacool2022. While OptiFit can effectively fit new sequence data to
existing OTU clusters, it is unknown if the use of OptiFit will have an
impact on classification performance. Here, we tested the ability of
OptiFit to cluster new sequence data into existing OTU clusters for the
purpose of classifying disease based on gut microbiome composition.

We compared the ability of several approaches for assigning 16S rRNA
gene sequences to OTUs including, *de novo* and reference-based clustering. For reference-based clustering, we used closed-reference clustering to a public database (database-reference-based) and to OTUs generated froma subset of the samples (self-reference-based). To test how the model performance
compared between these approaches, we used a publicly available dataset
of 16S rRNA gene sequences from stool samples of healthy subjects (n =
226) as well as subjects with screen-relevant neoplasia (SRN) consisting
of advanced adenoma and carcinoma (n = 229) @baxter2016. For the *de
novo* workflows, all the 16S rRNA sequence data was clustered into OTUs.
The OTU clustering was conducted using two common algorithms: 1) the
OptiClust algorithm in mothur [@westcott2017] and 2) the VSEARCH algorithm
used in QIIME2 [@rognes2016; @bolyen2019]. For both algorithms, the
resulting abundance data was then split into training and testing sets,
where the training set was used to tune hyperparameters and ultimately
train and select the model. The model was applied to the testing set and
performance evaluated (Figure 1A). We also conducted reference-based OTU
clustering using OptiFit to fit the sequence data into OTUs based on the
greengenes reference database. To compare with another commonly used
method, we also used the VSEARCH algorithm to fit the sequence data to
the greengenes reference (Figure 1B). In the OptiFit self-reference
workflow, the data was split into a training and a testing set. The
training set was clustered into OTUs and used to train a classification
model. The OptiFit algorithm was used to fit sequence data of samples
not part of the original dataset into the existing OTUs, and used the
same model to classify the samples (Figure 1C). For each of the
workflows the process was repeated for 100 random splits of the data to
account for variation caused by the choice of the random number
generator seed.

@@

We first examined the quality of the resulting OTU clusters from each
method using the Matthews correlation coefficient (MCC). MCC is a metric
used to measure OTU cluster quality based on the similarity of all pairs
of sequences and whether they are appropriately clustered or not
@westcott2015. MCC scores range between negative one and one, and
measure how well clustering assignment correlates with the distance
between sequences. To ensure that OptiFit appropriately integrated new
sequence data into the existing OTUs, we expected the MCC scores
produced by the OptiFit workflow to be similar to that of *de novo*
clustering using the OptiClust algorithm. In the OptiFit workflow the
test data was fit to the clustered training data for each of the 100
data splits resulting in an MCC score for each split of the data. In the
remaining workflows, the data was only clustered once and then split
into the training and testing sets resulting in a single MCC score for
each method. Indeed, the MCC scores were similar between the OptiClust
*de novo* (MCC = `r opticlust_denovo_mcc`) and OptiFit self-reference
workflows (average MCC = `r optifit_self_avg_mcc %>% pull(mean_mcc)`,
standard deviation = `r optifit_self_avg_mcc %>% pull(sd_mcc)`).
Consistent with prior findings, the reference-based methods produced
lower MCC scores (OptiFit Greengenes MCC = `r optifit_gg_mcc`; VSEARCH
Greengenes MCC = `r vsearch_gg_mcc`) than the *de novo* methods (OptiClust *de novo* MCC = `r opticlust_denovo_mcc`;
VSEARCH *de novo* MCC = `r vsearch_denovo_mcc`) @sovacool2022. Another metric we
examined for the OptiFit workflow was the fraction of sequences from the
test set that mapped to the reference OTUs. Since sequences that did not
map to reference OTUs were eliminated, if a high percentage of reads did
not map to an OTU we expected this loss of data to negatively impact
classification performance. We found that loss of data was not an issue
since on average
`r avg_frac_mapped %>% filter(methods == "optifit_self", set == "test") %>% pull(mean_frac) %>% round(digits=1)`%
(standard deviation =
`r avg_frac_mapped %>% filter(methods == "optifit_self", set == "test") %>% pull(sd_frac) %>% round(digits=2)`%)
of sequences in the test set mapped to the reference OTUs. This number
is higher than the average fraction of reads mapped in the OptiFit
Greengenes workflow (
`r avg_frac_mapped %>% filter(methods == "optifit_gg",set == "test") %>% pull(mean_frac) %>% round(digits=1)`%
+/-
`r avg_frac_mapped %>% filter(methods == "optifit_gg", set == "test") %>% pull(sd_frac) %>% round(digits=1)`).
These results indicate that the OptiFit self-reference method performed
as well as the OptiClust *de novo* method and better than using an
external database.

@@

get_pval <- function(pvals,met,g1,g2){
  pvals %>% 
    filter(metric == met) %>% 
    filter(group1 == g1 & group2 == g2 | group1 == g2 & group2 == g1) %>% 
    pull(p_value) %>% round(digits = 3)
}
```

We next assessed model performance using OTU relative abundances from
the training data from the workflows to train a model to predict SRNs
and used the model on the held out data. Using the predicted and actual
diagnosis classification, we calculated the area under the receiver
operating characteristic curve (AUROC) for each data split. During
cross-validation (CV) training, the performance of the OptFit self-reference and OptiClust *de novo* models were not significantly different (p-value =
`r get_pval(pvals, "cv_metric_AUC", "opticlust_denovo", "optifit_self")`;
Figure 2A), while performance for both VSEARCH methods was significantly lower
than the OptiClust *de novo*, OptiFit self, and OptiFit Greengenes
methods (p-values \< 0.05). The trained model was then applied to the
test data classifying samples as either control or SRN. Both VSEARCH
methods perform slightly worse than the OptiClust *de novo* method
(both p-values \< 0.05). However the performance on the test data for the OptiClust *de novo*, OptiFit Greengenes, and
OptiFit self-reference approaches were not significantly different (p-value \> 0.05; Figures 2B and 2C).
These results indicate that new data could be fit to existing OTU
clusters using OptiFit without impacting model performance.

>>>>

@@

There is increasing evidence for an association between the composition of the gut microbiome and a variety of diseases, such as crohn's disease and colorectal cancer [@Morgan2012;@Sobhani2011]. There is great potential to diagnose disease with gut microbiome sequence data and machine learning. Taxonomic composition of microbial
communities can be assessed using amplicon sequencing of the 16S rRNA
gene, which is the input to classification models. Analysis of 16S rRNA
gene sequence data generally relies on assigning sequences into
operational taxonomic units (OTUs). The process of OTU clustering can
either be reference-based or *de novo*. The quality of OTUs generated
with reference-based clustering is generally poor compared to those
generated with *de novo* clustering @westcott2015. While *de novo* clustering produces high-quality OTU clusters where sequences are accurately grouped based on similarity thresholds, the resulting OTU clusters depend on the sequences within the dataset and the addition of new data has the potential to redefine OTU cluster composition. The unstable nature of *de novo* OTU clustering complicates deployment of
machine learning models since integration of additional data requires
reclustering all the data and retraining the model. The ability to
integrate new data into a validated model without reclustering and
retraining could allow for the application of a single model that can
continually classify new data. Recently, Sovacool *et al.* introduced
OptiFit, a method for fitting new sequence data into existing OTUs
@sovacool2022. While OptiFit can effectively fit new sequence data to
existing OTU clusters, it is unknown if the use of OptiFit will have an
impact on classification performance. Here, we tested the ability of
OptiFit to cluster new sequence data into existing OTU clusters for the
purpose of classifying disease based on gut microbiome composition.

We compared the ability of several approaches for assigning 16S rRNA
gene sequences to OTUs including, *de novo* and reference-based clustering. For reference-based clustering, we used closed-reference clustering to a public database (database-reference-based) and to OTUs generated from a subset of the samples (self-reference-based). To test how the model performance
compared between these approaches, we used a publicly available dataset
of 16S rRNA gene sequences from stool samples of healthy subjects (n =
226) as well as subjects with screen-relevant neoplasia (SRN) consisting
of advanced adenoma and carcinoma (n = 229) @baxter2016. For the *de
novo* workflows, all the 16S rRNA sequence data was clustered into OTUs.
The OTU clustering was conducted using two common algorithms: 1) the
OptiClust algorithm in mothur [@westcott2017] and 2) the VSEARCH algorithm
used in QIIME2 [@rognes2016; @bolyen2019]. For both algorithms, the
resulting abundance data was then split into training and testing sets,
where the training set was used to tune hyperparameters and ultimately
train and select the model. The model was applied to the testing set and
performance evaluated (Figure 1A). We also conducted reference-based OTU
clustering using OptiFit to fit the sequence data into OTUs based on the
greengenes reference database. To compare with another commonly used
method, we also used the VSEARCH algorithm to fit the sequence data to
the greengenes reference (Figure 1B). In the OptiFit self-reference
workflow, the data was split into a training and a testing set. The
training set was clustered into OTUs and used to train a classification
model. The OptiFit algorithm was used to fit sequence data of samples
not part of the original dataset into the existing OTUs, and used the
same model to classify the samples (Figure 1C). For each of the
workflows the process was repeated for 100 random splits of the data to
account for variation caused by the choice of the random number
generator seed.

@@

We first examined the quality of the resulting OTU clusters from each
method using the Matthews correlation coefficient (MCC). MCC is a metric
used to measure OTU cluster quality based on the similarity of all pairs
of sequences and whether they are appropriately clustered or not
@westcott2015. MCC scores range between negative one and one, and
measure how well clustering assignment correlates with the distance
between sequences. To ensure that OptiFit appropriately integrated new
sequence data into the existing OTUs, we expected the MCC scores
produced by the OptiFit workflow to be similar to that of *de novo*
clustering using the OptiClust algorithm. In the OptiFit workflow the
test data was fit to the clustered training data for each of the 100
data splits resulting in an MCC score for each split of the data. In the
remaining workflows, the data was only clustered once and then split
into the training and testing sets resulting in a single MCC score for
each method. Indeed, the MCC scores were similar between the OptiClust
*de novo* (MCC = `r opticlust_denovo_mcc`) and OptiFit self-reference
workflows (average MCC = `r optifit_self_avg_mcc %>% pull(mean_mcc)`,
standard deviation = `r optifit_self_avg_mcc %>% pull(sd_mcc)`).
Consistent with prior findings, the reference-based methods produced
lower MCC scores (OptiFit Greengenes MCC = `r optifit_gg_mcc`; VSEARCH
Greengenes MCC = `r vsearch_gg_mcc`) than the *de novo* methods (OptiClust *de novo* MCC = `r opticlust_denovo_mcc`;
VSEARCH *de novo* MCC = `r vsearch_denovo_mcc`) @sovacool2022. Another metric we
examined for the OptiFit workflow was the fraction of sequences from the
test set that mapped to the reference OTUs. Since sequences that did not
map to reference OTUs were eliminated, if a high percentage of reads did
not map to an OTU we expected this loss of data to negatively impact
classification performance. We found that loss of data was not an issue
since on average
`r avg_frac_mapped %>% filter(methods == "optifit_self", set == "test") %>% pull(mean_frac) %>% round(digits=1)`%
(standard deviation =
`r avg_frac_mapped %>% filter(methods == "optifit_self", set == "test") %>% pull(sd_frac) %>% round(digits=2)`%)
of sequences in the subsampled test set mapped to the reference OTUs. This number
is higher than the average fraction of reads mapped in the OptiFit
Greengenes workflow (
`r avg_frac_mapped %>% filter(methods == "optifit_gg",set == "test") %>% pull(mean_frac) %>% round(digits=1)`%
+/-
`r avg_frac_mapped %>% filter(methods == "optifit_gg", set == "test") %>% pull(sd_frac) %>% round(digits=1)`).
These results indicate that the OptiFit self-reference method performed
as well as the OptiClust *de novo* method and better than using an
external database.

@@

get_pval <- function(pvals,met,g1,g2){
  val <- pvals %>% 
    filter(metric == met) %>% 
    filter(group1 == g1 & group2 == g2 | group1 == g2 & group2 == g1) %>% 
    pull(p_value)

  format(round(val,digits = 3),nsmall=3)
}
#get_pval(pvals,"AUC","opticlust_denovo","vsearch_gg")
```

We next assessed model performance using OTU relative abundances from
the training data from the workflows to train a model to predict SRNs
and used the model on the held out data. Using the predicted and actual
diagnosis classification, we calculated the area under the receiver
operating characteristic curve (AUROC) for each data split. During
cross-validation (CV) training, the performance of the OptiFit self-reference and OptiClust *de novo* models were not significantly different (p-value =
`r get_pval(pvals, "cv_metric_AUC", "opticlust_denovo", "optifit_self")`;
Figure 2A), while performance for both VSEARCH methods was significantly lower
than the OptiClust *de novo*, OptiFit self, and OptiFit Greengenes
methods (p-values \< 0.05). The trained model was then applied to the
test data classifying samples as either control or SRN. The VSEARCH Greengenes method performed slightly worse than the OptiClust *de novo* method (p-value = `r get_pval(pvals,"AUC","opticlust_denovo","vsearch_gg")`). However the performance on the test data for the OptiClust *de novo*, OptiFit Greengenes, OptiFit self-reference, and VSEARCH *de novo* approaches were not significantly different (p-values \> 0.05; Figures 2B and 2C).
These results indicate that new data could be fit to existing OTU
clusters using OptiFit without impacting model performance.

### 2023-04-27 submission/manuscript.tex
final revisions

\textbf{observation format} (max 1200 words, 2 figures, 25 ref)

@@

There is increasing evidence for an association between the composition
of the gut microbiome and a variety of diseases, such as crohn's disease
and colorectal cancer (\protect\hyperlink{ref-Morgan2012}{1},
\protect\hyperlink{ref-Sobhani2011}{2}). There is great potential
diagnose of disease with gut microbiome sequence data and machine
learning. Taxonomic composition of microbial communities can be assessed
using amplicon sequencing of the 16S rRNA gene, which is the input to
classification models. Analysis of 16S rRNA gene sequence data generally
relies on assigning sequences into operational taxonomic units (OTUs).
The process of OTU clustering can either be reference-based or \emph{de
novo}. The quality of OTUs generated with reference-based clustering is
generally poor compared to those generated with \emph{de novo}
clustering (\protect\hyperlink{ref-westcott2015}{3}). While \emph{de
novo} clustering produces high-quality OTU clusters where sequences are
accurately grouped based on similarity thresholds, the resulting OTU
clusters depend on the sequences within the dataset and the addition of
new data has the potential to redefine OTU cluster composition. The
unstable nature of \emph{de novo} OTU clustering complicates deployment
of machine learning models since integration of additional data requires
reclustering all the data and retraining the model. The ability to
integrate new data into a validated model without reclustering and
retraining could allow for the application of a single model that can
continually classify new data. Recently, Sovacool \emph{et al.}
introduced OptiFit, a method for fitting new sequence data into existing
OTUs (\protect\hyperlink{ref-sovacool2022}{4}). While OptiFit can
effectively fit new sequence data to existing OTU clusters, it is
unknown if the use of OptiFit will have an impact on classification
performance. Here, we tested the ability of OptiFit to cluster new
sequence data into existing OTU clusters for the purpose of classifying
disease based on gut microbiome composition.

We compared the ability of several approaches for assigning 16S rRNA
gene sequences to OTUs including, \emph{de novo} and reference-based
clustering. For reference-based clustering, we used closed-reference
clustering to a public database (database-reference-based) and to OTUs
generated froma subset of the samples (self-reference-based). To test
how the model performance compared between these approaches, we used a
publicly available dataset of 16S rRNA gene sequences from stool samples
of healthy subjects (n = 226) as well as subjects with screen-relevant
neoplasia (SRN) consisting of advanced adenoma and carcinoma (n = 229)
(\protect\hyperlink{ref-baxter2016}{5}). For the \emph{de novo}
workflows, all the 16S rRNA sequence data was clustered into OTUs. The
OTU clustering was conducted using two common algorithms: 1) the
OptiClust algorithm in mothur (\protect\hyperlink{ref-westcott2017}{6})
and 2) the VSEARCH algorithm used in QIIME2
(\protect\hyperlink{ref-rognes2016}{7},
\protect\hyperlink{ref-bolyen2019}{8}). For both algorithms, the
resulting abundance data was then split into training and testing sets,
where the training set was used to tune hyperparameters and ultimately
train and select the model. The model was applied to the testing set and
performance evaluated (Figure 1A). We also conducted reference-based OTU
clustering using OptiFit to fit the sequence data into OTUs based on the
greengenes reference database. To compare with another commonly used
method, we also used the VSEARCH algorithm to fit the sequence data to
the greengenes reference (Figure 1B). In the OptiFit self-reference
workflow, the data was split into a training and a testing set. The
training set was clustered into OTUs and used to train a classification
model. The OptiFit algorithm was used to fit sequence data of samples
not part of the original dataset into the existing OTUs, and used the
same model to classify the samples (Figure 1C). For each of the
workflows the process was repeated for 100 random splits of the data to
account for variation caused by the choice of the random number
generator seed.

We first examined the quality of the resulting OTU clusters from each
method using the Matthews correlation coefficient (MCC). MCC is a metric
used to measure OTU cluster quality based on the similarity of all pairs
of sequences and whether they are appropriately clustered or not
(\protect\hyperlink{ref-westcott2015}{3}). MCC scores range between
negative one and one, and measure how well clustering assignment
correlates with the distance between sequences. To ensure that OptiFit
appropriately integrated new sequence data into the existing OTUs, we
expected the MCC scores produced by the OptiFit workflow to be similar
to that of \emph{de novo} clustering using the OptiClust algorithm. In
the OptiFit workflow the test data was fit to the clustered training
data for each of the 100 data splits resulting in an MCC score for each
split of the data. In the remaining workflows, the data was only
clustered once and then split into the training and testing sets
resulting in a single MCC score for each method. Indeed, the MCC scores
were similar between the OptiClust \emph{de novo} (MCC = 0.884) and
OptiFit self-reference workflows (average MCC = 0.879, standard
deviation = 0.002). Consistent with prior findings, the reference-based
methods produced lower MCC scores (OptiFit Greengenes MCC = 0.786;
VSEARCH Greengenes MCC = 0.531) than the \emph{de novo} methods
(OptiClust \emph{de novo} MCC = 0.884; VSEARCH \emph{de novo} MCC =
0.641) (\protect\hyperlink{ref-sovacool2022}{4}). Another metric we
examined for the OptiFit workflow was the fraction of sequences from the
test set that mapped to the reference OTUs. Since sequences that did not
map to reference OTUs were eliminated, if a high percentage of reads did
not map to an OTU we expected this loss of data to negatively impact
classification performance. We found that loss of data was not an issue
since on average 99.8\% (standard deviation = 0.68\%) of sequences in
the test set mapped to the reference OTUs. This number is higher than
the average fraction of reads mapped in the OptiFit Greengenes workflow
( 96.8\% +/- 3.5). These results indicate that the OptiFit
self-reference method performed as well as the OptiClust \emph{de novo}
method and better than using an external database.

We next assessed model performance using OTU relative abundances from
the training data from the workflows to train a model to predict SRNs
and used the model on the held out data. Using the predicted and actual
diagnosis classification, we calculated the area under the receiver
operating characteristic curve (AUROC) for each data split. During
cross-validation (CV) training, the performance of the OptFit
self-reference and OptiClust \emph{de novo} models were not
significantly different (p-value = 0.071; Figure 2A), while performance
for both VSEARCH methods was significantly lower than the OptiClust
\emph{de novo}, OptiFit self, and OptiFit Greengenes methods (p-values
\textless{} 0.05). The trained model was then applied to the test data
classifying samples as either control or SRN. Both VSEARCH methods
perform slightly worse than the OptiClust \emph{de novo} method (both
p-values \textless{} 0.05). However the performance on the test data for
the OptiClust \emph{de novo}, OptiFit Greengenes, and OptiFit
self-reference approaches were not significantly different (p-value
\textgreater{} 0.05; Figures 2B and 2C). These results indicate that new
data could be fit to existing OTU clusters using OptiFit without
impacting model performance.

>>>>

@@

There is increasing evidence for an association between the composition
of the gut microbiome and a variety of diseases, such as crohn's disease
and colorectal cancer (\protect\hyperlink{ref-Morgan2012}{1},
\protect\hyperlink{ref-Sobhani2011}{2}). There is great potential to
diagnose disease with gut microbiome sequence data and machine learning.
Taxonomic composition of microbial communities can be assessed using
amplicon sequencing of the 16S rRNA gene, which is the input to
classification models. Analysis of 16S rRNA gene sequence data generally
relies on assigning sequences into operational taxonomic units (OTUs).
The process of OTU clustering can either be reference-based or \emph{de
novo}. The quality of OTUs generated with reference-based clustering is
generally poor compared to those generated with \emph{de novo}
clustering (\protect\hyperlink{ref-westcott2015}{3}). While \emph{de
novo} clustering produces high-quality OTU clusters where sequences are
accurately grouped based on similarity thresholds, the resulting OTU
clusters depend on the sequences within the dataset and the addition of
new data has the potential to redefine OTU cluster composition. The
unstable nature of \emph{de novo} OTU clustering complicates deployment
of machine learning models since integration of additional data requires
reclustering all the data and retraining the model. The ability to
integrate new data into a validated model without reclustering and
retraining could allow for the application of a single model that can
continually classify new data. Recently, Sovacool \emph{et al.}
introduced OptiFit, a method for fitting new sequence data into existing
OTUs (\protect\hyperlink{ref-sovacool2022}{4}). While OptiFit can
effectively fit new sequence data to existing OTU clusters, it is
unknown if the use of OptiFit will have an impact on classification
performance. Here, we tested the ability of OptiFit to cluster new
sequence data into existing OTU clusters for the purpose of classifying
disease based on gut microbiome composition.

We compared the ability of several approaches for assigning 16S rRNA
gene sequences to OTUs including, \emph{de novo} and reference-based
clustering. For reference-based clustering, we used closed-reference
clustering to a public database (database-reference-based) and to OTUs
generated from a subset of the samples (self-reference-based). To test
how the model performance compared between these approaches, we used a
publicly available dataset of 16S rRNA gene sequences from stool samples
of healthy subjects (n = 226) as well as subjects with screen-relevant
neoplasia (SRN) consisting of advanced adenoma and carcinoma (n = 229)
(\protect\hyperlink{ref-baxter2016}{5}). For the \emph{de novo}
workflows, all the 16S rRNA sequence data was clustered into OTUs. The
OTU clustering was conducted using two common algorithms: 1) the
OptiClust algorithm in mothur (\protect\hyperlink{ref-westcott2017}{6})
and 2) the VSEARCH algorithm used in QIIME2
(\protect\hyperlink{ref-rognes2016}{7},
\protect\hyperlink{ref-bolyen2019}{8}). For both algorithms, the
resulting abundance data was then split into training and testing sets,
where the training set was used to tune hyperparameters and ultimately
train and select the model. The model was applied to the testing set and
performance evaluated (Figure 1A). We also conducted reference-based OTU
clustering using OptiFit to fit the sequence data into OTUs based on the
greengenes reference database. To compare with another commonly used
method, we also used the VSEARCH algorithm to fit the sequence data to
the greengenes reference (Figure 1B). In the OptiFit self-reference
workflow, the data was split into a training and a testing set. The
training set was clustered into OTUs and used to train a classification
model. The OptiFit algorithm was used to fit sequence data of samples
not part of the original dataset into the existing OTUs, and used the
same model to classify the samples (Figure 1C). For each of the
workflows the process was repeated for 100 random splits of the data to
account for variation caused by the choice of the random number
generator seed.

We first examined the quality of the resulting OTU clusters from each
method using the Matthews correlation coefficient (MCC). MCC is a metric
used to measure OTU cluster quality based on the similarity of all pairs
of sequences and whether they are appropriately clustered or not
(\protect\hyperlink{ref-westcott2015}{3}). MCC scores range between
negative one and one, and measure how well clustering assignment
correlates with the distance between sequences. To ensure that OptiFit
appropriately integrated new sequence data into the existing OTUs, we
expected the MCC scores produced by the OptiFit workflow to be similar
to that of \emph{de novo} clustering using the OptiClust algorithm. In
the OptiFit workflow the test data was fit to the clustered training
data for each of the 100 data splits resulting in an MCC score for each
split of the data. In the remaining workflows, the data was only
clustered once and then split into the training and testing sets
resulting in a single MCC score for each method. Indeed, the MCC scores
were similar between the OptiClust \emph{de novo} (MCC = 0.884) and
OptiFit self-reference workflows (average MCC = 0.879, standard
deviation = 0.002). Consistent with prior findings, the reference-based
methods produced lower MCC scores (OptiFit Greengenes MCC = 0.786;
VSEARCH Greengenes MCC = 0.531) than the \emph{de novo} methods
(OptiClust \emph{de novo} MCC = 0.884; VSEARCH \emph{de novo} MCC =
0.641) (\protect\hyperlink{ref-sovacool2022}{4}). Another metric we
examined for the OptiFit workflow was the fraction of sequences from the
test set that mapped to the reference OTUs. Since sequences that did not
map to reference OTUs were eliminated, if a high percentage of reads did
not map to an OTU we expected this loss of data to negatively impact
classification performance. We found that loss of data was not an issue
since on average 99.8\% (standard deviation = 0.68\%) of sequences in
the subsampled test set mapped to the reference OTUs. This number is
higher than the average fraction of reads mapped in the OptiFit
Greengenes workflow ( 96.8\% +/- 3.5). These results indicate that the
OptiFit self-reference method performed as well as the OptiClust
\emph{de novo} method and better than using an external database.

We next assessed model performance using OTU relative abundances from
the training data from the workflows to train a model to predict SRNs
and used the model on the held out data. Using the predicted and actual
diagnosis classification, we calculated the area under the receiver
operating characteristic curve (AUROC) for each data split. During
cross-validation (CV) training, the performance of the OptiFit
self-reference and OptiClust \emph{de novo} models were not
significantly different (p-value = 0.066; Figure 2A), while performance
for both VSEARCH methods was significantly lower than the OptiClust
\emph{de novo}, OptiFit self, and OptiFit Greengenes methods (p-values
\textless{} 0.05). The trained model was then applied to the test data
classifying samples as either control or SRN. The VSEARCH Greengenes
method performed slightly worse than the OptiClust \emph{de novo} method
(p-value = 0.030). However the performance on the test data for the
OptiClust \emph{de novo}, OptiFit Greengenes, OptiFit self-reference,
and VSEARCH \emph{de novo} approaches were not significantly different
(p-values \textgreater{} 0.05; Figures 2B and 2C). These results
indicate that new data could be fit to existing OTU clusters using
OptiFit without impacting model performance.
